Transition path sampling based calculations of free energies of enzymatic reactions: the case of human MAT2A and plasmodium vivax adenosine deaminase

### 2022-04-20 tex/manuscript.tex
added some more text and figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article,layout=twocolumn]{achemso}

@@

simulations. For the MAT2A enzyme the substrates were modified to 
L-methionine and adenosyl triphosphate (ATP) whereas for the \textit{pv}ADA

@@

using potassium ions. Both the systems were minimized with 50 steps of the

@@

equilibration to prepare the systems for production runs.  

@@

We used TPS to generate an ensemble of reactive trajectories. 
The time step that we used for all our calculations was $1\;fs$ with a total
of 250 steps forward and 250 steps backward which amounts to a total of $0.5\;ps$
trajectory.   
\subsubsection{human MAT2A}
The key bond parameters for the reaction in this system consists of the 
oxygen atom on the ATP, the carbon atom on the adenosyl group and the 
sulphur atom on the L-methionine. The reactant basin is defined by the 
oxygen-carbon bond distance being less than 1.7 {\AA} and the product basin 
is characterized by the suphur-carbon bond distance being less than 
1.7 {\AA}. The initial reactive trajectory was generated by the application 
of a bias harmonic force between the oxygen and the carbon atoms 
and running dynamics for $250\;fs$. The shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. Then with the new momenta
the system is propagated forward and backward in time for $250\;fs$.
A total of 120 reactive trajectories were generated.

\subsubsection{plasmodium vivax ADA}
The $pv$ADA reaction is 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
For the free energy calculations we divided the reaction coordinate into 13 windows for 
MAT2A and 20 windows for pvADA with a $15\%$ overlap on each side of the window.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article,layout=traditional]{achemso}

@@

simulations. For calculations involving the MAT2A enzyme the substrates were modified to 
L-methionine and adenosyl triphosphate (ATP) whereas for calculations with the \textit{pv}ADA

@@

using potassium ions. 
Both the systems were minimized with 50 steps of the

@@

equilibration to prepare the systems for TPS simulations.  

@@

TPS calculations collect reactive trajectories based on the Monte Carlo algorithm 
and requires a proper definition of the reactant and product states.

The reaction catalyzed by the human MAT2A enzyme is shown in Fig. \ref{fig:mat2a-reaction}
between adenosyl $5'$-triphosphate (ATP) and L-methionine (MET) molecules resulting in the formation of
S-adenosyl-L-methionine (SAM). SAM is an essential metabolite which is distributed to almost all 
body tissues and fluids, is an universal methyl donor and is of
fundamental importance to metabolism of compounds such as hormones, neureotransmitters, proteins, 
and nucleic acids. \cite{Friedel89Drugs38p389} The reaction was experimentally characterized by 
Firestone et al. \cite{Firestone17JAmChemSoc139p13754} to follow a S$_{\text{N}}2$ transition 
state. The key bond parameters for this reaction consists of the distance between the
nucleophilic sulfur atom of MET and the $5'-C$ atom of the ATP ($S-C$), and the distance between 
oxygen atom of the phosphate group and the $5'-C$ atom of ATP ($O-C$).
For the TPS calculations we define the reactant state to have $O-C$ distance to be 
less than 1.7 {\AA} and the product state to have the $S-C$ distance being less than 1.7 {\AA}.

\begin{figure}
\includegraphics[scale=0.6]{figures/mat2a-reaction.png}
\caption{The reaction catalyzed by human MAT2A enzyme}
\label{fig:mat2a-reaction}
\end{figure}

The reaction catalyzed by $pv$ADA is shown in  
Fig. \ref{fig:ada-reaction} which is between the adenosine molecule and a hydroxide ion
which results in the formation of inosine with ammonia as a by product. 
$pv$ADA is an essential enzyme in the purine salvage pathway for plasmodium vivax and 
inhibiting this enzyme is a promising approach to treat malaria. \cite{Madrid08JBiolChem283p35899}  
This reaction proceeds with the nucleophilic O atom (O$^{\text{hydroxyl}}$) of the hydroxide anion
attacking the electrophilic C6 atom of the adenosyl ring leading to the formation of a Meisenheimer 
complex while the N1 atom on the adenosyl ring starts to get 
protonated by the Glu229 residue of $pv$ADA. At the transition state, experimental findings 
suggest that the dissociation of the C6$-$N6 bond as well as the protonation of the N1 
atom are almost complete. \cite{Luo07JAmChemSoc129p8008} The key bond parameters are 
the distance between the approaching hydroxyl oxygen and the C6 atom ($O-C$) and the 
dissociating C6$-$N6 ($N-C$) bond distance. For the TPS simulaions we defined the reactant state to 
have the $O-C$ distance being less than $1.7$ {\AA} and the product state to have the $O-C$ 
distance being less than $1.7$ {\AA}.

\begin{figure}
\includegraphics[scale=0.6]{figures/ada-new.png}
\caption{The reaction catalyzed by adenosine deaminase enzyme}
\label{fig:ada-reaction}
\end{figure}

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. 
The time step that we used for all our calculations was $1\;fs$ and we propagate the 
dynamics for a total of 250 steps forward and 250 steps backward which amounts to a 
total of $0.5\;ps$. 
A total of 150 and 100 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively.

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
Window based free energy calculations such as WHAM divide the reaction coordinate
into windows and calculate the individual populations within the windows and 
decouple the applied bias with some formulas. 
For the free energy calculations we divided the reaction coordinate into 13 windows for 
MAT2A and 20 windows for pvADA with a $15\%$ overlap on each side of the window.
